A Formal Treatment of Bidirectional Typing

### 2024-01-18 tex/ESOP24/BiSig.tex
[ tex ] Fight with the page limit
[ tex ] Include the running head
Sprinkled the paper with commas; miscellaneous revisions

\documentclass[orivec,envcountsect,envcountsame]{llncs}

@@

\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
\institute{Institute of Information Science, Academia Sinica, Taiwan}
\title{A formal treatment of bidirectional typing}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a \emph{proof-relevant} type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a raw term.
The entire theory is formally implemented in \Agda, so as a by-product of our formalism we provide a \emph{verified} generator of proof-relevant type synthesisers.
\end{abstract}

@@

\subfile{S7-formalisation}
\input{appendix}

>>>>

\documentclass[runningheads,orivec,envcountsect,envcountsame]{llncs}

@@

\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
\institute{Institute of Information Science, Academia Sinica, Taipei, Taiwan}
\title{A Formal Treatment of Bidirectional Typing}
\authorrunning{L.-T.~Chen and H.-S.~Ko}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a \emph{proof-relevant} type synthesiser which, given an input abstract syntax tree, constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness are studied universally for all signatures.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations.
The entire theory is formally implemented in \Agda, so we provide a \emph{verified} generator of proof-relevant type synthesisers as a by-product of our formalism.
\end{abstract}

% Due to ESOP's page limit, the appendix is only available in the ArXiv version.

\newif\ifarxiv

@@

\ifarxiv
\subfile{S7-formalisation}

\input{appendix}
\fi

### 2024-01-18 tex/ESOP24/S1-intro.tex
Sprinkled the paper with commas; miscellaneous revisions

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism entails undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, deciding the types of raw terms that meet some syntactic criteria and typically contain annotations.
In their survey paper, \varcitet{Dunfield2021}{ summarised} the principles of bidirectional type synthesis and its wide coverage of languages with simple, polymorphic, dependent, and gradual types, among others.

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are input.
The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis.
%this situation is in stark contrast to parsing, for which there are general theories and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
%While it is usually straightforward to derive a bidirectional type synthesis algorithm,
\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
%Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
In this paper, we develop such a theory of bidirectional typing with the proof assistant \Agda.
%and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
%In particular, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

@@

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one: traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
%\Josh{The algorithm adds annotations to the input term?}
can be transformed to a well-typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
Such an algorithm is accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates algorithmic soundness, completeness, and decidability \emph{in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

@@

As we mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} to explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
Thus, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide if the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a proof-relevant \emph{mode decorator}, which either
\begin{inlineenum}
  \item construct a mode derivation for a raw term, or
  \item provides information that refutes the existence of any mode derivation and pinpoints missing annotations.
\end{inlineenum}
Then a bidirectional type synthesiser is only required to decide the typability of mode-decorated raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
As illustrated by \citet{Dunfield2021}, it seems that the implications of mode-correctness have only been addressed informally so far and mode-correctness is not yet formally defined as a \emph{property of languages}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed~\cite{Chen2024a} with \Agda but translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, with no gap in between.

@@

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example} and further developments are discussed in \cref{sec:future}.
\ifarxiv
A demonstration is given in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.
\fi

>>>>

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism entails undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, deciding the types of raw terms that meet some syntactic criteria and typically contain annotations.
In their survey paper~\cite{Dunfield2021}, Dunfield and Krishnaswami summarised the principles of bidirectional type synthesis and its wide coverage of languages with simple, polymorphic, dependent, and gradual types, among others.

While type inference is not decidable in general, for certain kinds of terms it is still possible to synthesise their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{synthesis} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are input.
The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis.
%this situation is in stark contrast to parsing, for which there are general theories and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
%While it is usually straightforward to derive a bidirectional type synthesis algorithm,
Dunfield and Krishnaswami present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
%Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
In this paper, we develop such a theory of bidirectional typing with the proof assistant \Agda.
%and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
%In particular, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

@@

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one: traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
%\Josh{The algorithm adds annotations to the input term?}
can be transformed to a well-typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
Such an algorithm is accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term, and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory, and formulates algorithmic soundness, completeness, and decidability \emph{in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

@@

As we mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} to explicitly take annotations into account, and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
Thus, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide if the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a proof-relevant \emph{mode decorator}, which either
\begin{inlineenum}
  \item construct a mode derivation for a raw term, or
  \item provides information that refutes the existence of any mode derivation and pinpoints missing annotations.
\end{inlineenum}
Then a bidirectional type synthesiser is only required to decide the typability of mode-decorated raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term.
Our completeness is simpler and more useful than annotatability, which is a typical formulation of completeness in the literature of bidirectional typing~\cite[Section~3.2]{Dunfield2021}.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
As illustrated by \citet{Dunfield2021}, it seems that the implications of mode-correctness have only been addressed informally so far, and mode-correctness is not yet formally defined as a \emph{property of languages}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed with \Agda, but is translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, which is available publicly on Zenodo~\cite{Chen2024a}.

@@

This paper is structured as follows.
We present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}, and further developments are discussed in \cref{sec:future}.
\ifarxiv
A demonstration is given in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.
\fi

### 2024-01-18 tex/ESOP24/S2-outline.tex
[ tex ] Minor changes
[ tex ] Fix minor issues
Sprinkled the paper with commas; miscellaneous revisions

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term as input, and produce a typing derivation for the term if possible.
To give more precise definitions: the raw terms for simply typed $\lambda$-calculus are defined%
\footnote{The usual condition about named representations of variables is omitted.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.

@@

Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of the simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} has an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.

@@

\begin{defn} \label{def:STLC-mode-decorator}
A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ if $V \vdash \isTerm{t}^{\syn}$.
\end{defn}

>>>>

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term as input, and produce a typing derivation for the term if possible.
To give more precise definitions: the raw terms for simply typed $\lambda$-calculus are defined%
\footnote{The usual conditions about named representations of variables are omitted.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.

@@

Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} has an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.

@@

\begin{defn} \label{def:STLC-mode-decorator}
  A \emph{mode decorator} decides for a raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
\end{defn}
